---
title: "Alpha Diversity"
author: "Katherine Mueller"
date: "3/9/2021"
output: html_document
---

###Initial Alpha Diversity Comparisons

##Setup
#Load Libraries
```{r}
library(readr)
library(phyloseq)
library(tibble)
library(dplyr)
library(ggplot2)
```

# Define Paths
```{r}
plastics.rds = "/sharedspace/PlasticPeople/plastics.rds"
```

# Check Phyloseq RDS
Load your phyloseq RDS and `print` it just to be sure things look sane
```{r}
plastics.ps = read_rds(plastics.rds)
print(plastics.ps)
```

```{r}
sample_variables(plastics.ps)
```


## Analysis
#Week One
Week one represents the time that samples were collected, before growth in individual conditions. Therefore, we would expect that the four samples for this timepoint should have the same taxa within a given location

```{r}
weekone.ps <- plastics.ps %>%
     subset_samples(Week == 1)
print(weekone.ps)
```

```{r}
plot_richness(weekone.ps, x = "Condition",
              measures = c("Observed"), color = "Condition") +
     geom_point() +
     facet_grid(~Location) +
     theme_bw() +
     labs(title = "Week 1 Observed ASVs")
```

```{r}
plot_richness(weekone.ps, x = "Condition",
              measures = c("Shannon"), color = "Condition") +
     geom_point() +
     facet_grid(~Location) +
     theme_bw() +
     labs(title = "Week 1 Shannon Diversity")
```