---
title: "Beta diversity"
author: "Brianna Petrone"
date: "3/31/2021"
output: html_document
---

# Setup

```{r}
# Install packages

# If not yet installed, need to install as follows
# install.packages("cowplot") # For multi-panel plot layouts
# devtools::install_github("jsilve24/driver") # For CLR transform
```

```{r include=FALSE, echo=FALSE}
# Load libraries
library(cowplot)
library(driver)
library(phyloseq)
library(RColorBrewer) # For plot colors
library(tidyverse) # For data wrangling tools
```

```{r}
# Set paths
data.dir <- '/data/project_data/argonne_data' # Data
scratch.dir <- '/home/guest/scratch/plastics' # Scratch
group.dir <- '/sharedspace/PlasticPeople' # Group
figs.dir <- file.path(scratch.dir, 'figures') # Figures
```

# Read in data

```{r}
# Phyloseq object
ps <- readRDS(file.path(group.dir, 'plastics.rds'))
ps
```

# Preprocessing

## Refactor sample data

To facilitate plotting

```{r}
sample_data(ps)$Location <- factor(sample_data(ps)$Location,
                                   labels = c('Activated sludge',
                                              'Elizabeth River',
                                              'Landfill',
                                              'Recycling center'))

sample_data(ps)$Environment <- 
     ifelse(sample_data(ps)$Location == 'Elizabeth River',
            yes = 1,
            no = 0) %>% 
     factor(labels = c('Waste management facilities', 
                       'Natural environment'))

sample_data(ps)$Week <- as.numeric(sample_data(ps)$Week)

sample_data(ps)$Condition <- factor(sample_data(ps)$Condition,
                                   labels = c('No treatment',
                                              'Polymer',
                                              'Polymer and monomer',
                                              'Monomer'))

# Add extra column for monomer present
sample_data(ps)$Monomer <- 
     ifelse(grepl('[Mm]onomer', sample_data(ps)$Condition),
            yes = 1,
            no = 0) %>% 
     factor(labels = c('Without styrene', 
                       'With styrene'))
```


## Transform

```{r}
# Plan ahead
# Want to add pseudocount before CLR-transforming; easiest to do this now, while data are still interpreted as counts

# Pseudocount of 1 for compatibility with log transform
otu_table(ps) <- otu_table(ps) + 1 

# Check it worked
asvtab <- otu_table(ps)@.Data
min(asvtab)
```

```{r}
# Transform to relative abundance
ps.ra <- transform_sample_counts(ps, function(x){x/sum(x)})
```

## Filter

Want to do a thoughtful filtering step here.  For streamgraphs tried glomming to family level, then only keeping taxa with an abundance of 1% in at least 4 samples (because we might expect each location to be reasonably different, but is tested in 4 conditions at each timepoint).  

See how stringent this is, and how many more taxa are included if relaxed (perhaps to 2 samples)?

At the ASV level, we may want to relax this further (say to 0.01%)?

```{r}
# 0.01% in at least 4 samples
min_fraction = 0.0001
min_samples = 4
filter_taxa(ps.ra, 
            function(x) sum(x >= min_fraction) >= min_samples, prune=TRUE)
```

This reduces the number of taxa by `r 100*(1-(3547/ntaxa(ps)))`%.

```{r}
# 0.01% in at least 2 samples
min_fraction = 0.0001
min_samples = 2
filter_taxa(ps.ra, 
            function(x) sum(x >= min_fraction) >= min_samples, prune=TRUE)
```

This reduces the number of taxa by `r 100*(1-(6069/ntaxa(ps)))`%.

```{r}
# 0.1% in at least 1 samples
min_fraction = 0.001
min_samples = 1
ps.ra.filt <- 
     filter_taxa(ps.ra, 
                 function(x) sum(x >= min_fraction) >= min_samples, prune=TRUE)
ps.ra.filt
```

This reduces the number of taxa by `r 100*(1-(2933/ntaxa(ps)))`%.

Choosing the third filter condition for now.  I think a tenth of a percent is a meaningful relative abundance, and I like permitting keeping the threshold at that amount in 1 sample because there's so little replication in this dataset.

# Ordinate

## By ASV

### PCA

#### CLR transform input

```{r}
# Pseudocount has already been added, pre-transformation, so no need to do here

# Get ASV table
asvtab <- otu_table(ps.ra.filt)@.Data

# Transform
asvtab.clr <- clr(asvtab)

# Replace names, but remember these aren't interpretable as strict ASVs any longer
colnames(asvtab.clr) <- colnames(asvtab)

# Make updated phyloseq object
ps.clr <- phyloseq(otu_table(asvtab.clr, taxa_are_rows = FALSE),
                   sample_data(sample_data(ps)),
                   tax_table(tax_table(ps)))
```

#### Run

```{r}
# Run PCA
pca <- prcomp(asvtab.clr)
```

#### Plot

```{r}
# Get sample data
samdf <- 
    sample_data(ps.clr) %>% 
    as(., 'data.frame') 

# Add back sample data to PCA results
pca.df <- 
     data.frame(pca$x) %>% 
     rownames_to_column(var = 'sample')
pca.df <- left_join(pca.df, rownames_to_column(samdf, var = 'sample'))

# Figure out axis limits to customize scale of data
# Find the largest magnitude datapoint in the first 2 PCs, then add 5% to it for 
# plotting "room"
limit <- max(abs(pca.df[, c('PC1','PC2')])) +
     0.05*(max(abs(pca.df[, c('PC1','PC2')])))

# % variance explained
eigs <- pca$sdev^2
ve.pc1 <- as.character(100*round(eigs[1] / sum(eigs), 3))
ve.pc2 <- as.character(100*round(eigs[2] / sum(eigs), 3))
```


```{r}
# Plot by location
# pca.plot <- 
#      ggplot(pca.df, aes(PC1, PC2, color = Environment)) +
#      geom_point() +
#      coord_equal() +
#      xlim(-limit, limit) + ylim(-limit, limit) + 
#      labs(x = paste0(' PC1 (', ve.pc1, '%)'),
#           y = paste0(' PC1 (', ve.pc2, '%)'),
#           color = 'Sample source') + 
#      scale_color_manual(values = c('gray60', 'skyblue3')) +
#      theme_bw()

a <-
     ggplot(pca.df, aes(PC1, PC2, color = Location)) +
     geom_point() +
     coord_equal() +
     xlim(-limit, limit) + ylim(-limit, limit) +
     labs(x = paste0(' Principal Component 1 (', ve.pc1, '%)'),
          y = paste0(' Principal Component 2 (', ve.pc2, '%)'),
          title = 'Sample source') +
     scale_color_manual(values = c('gray60', 'skyblue3', 'gray40', 'gray80')) +
     theme(legend.position = c(0.8, 0.8),
           plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"),
           panel.border = element_blank(), 
           panel.grid.major = element_blank(), 
           panel.grid.minor = element_blank(), 
           axis.line = element_line(colour = "black", 
                                    size = rel(1)), 
           legend.key = element_blank(), 
           legend.title = element_blank(), 
           panel.background = element_rect(fill = "white", 
                                           colour = NA),
           strip.background = element_rect(fill = "white", colour = "black", 
                                           size = rel(2)), complete = TRUE)

a
```

```{r}
# Plot by timecourse
b <- 
    ggplot(pca.df, aes(PC1, PC2, color = Week)) +
    geom_point() +
    coord_equal() +
    xlim(-limit, limit) + ylim(-limit, limit) + 
    labs(x = paste0(' PC 1 (', ve.pc1, '%)'),
         y = paste0(' PC 2 (', ve.pc2, '%)'),
         title = 'Bioreactor week') + 
    scale_color_distiller(palette = 'Greys', direction = 1) +
    theme(legend.position = c(0.8, 0.8),
          plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"),
          axis.title.y = element_blank(),
           panel.border = element_blank(), 
           panel.grid.major = element_blank(), 
           panel.grid.minor = element_blank(), 
           axis.line = element_line(colour = "black", 
                                    size = rel(1)), 
          legend.title = element_blank(), 
           legend.key = element_blank(), 
           panel.background = element_rect(fill = "white", 
                                           colour = NA),
           strip.background = element_rect(fill = "white", colour = "black", 
                                           size = rel(2)), complete = TRUE)


b
```

```{r}
# Plot by treatment
c <- 
     ggplot(pca.df, aes(PC1, PC2, color = Monomer)) +
     geom_point() +
     coord_equal() +
     xlim(-limit, limit) + ylim(-limit, limit) + 
     labs(x = paste0(' PC 1 (', ve.pc1, '%)'),
          y = paste0(' PC 2 (', ve.pc2, '%)'),
          title = 'Treatment condition') + 
     scale_color_manual(values = c('gray60', 'red3')) +
     theme(legend.position = c(0.8, 0.8),
           plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"),
           axis.title.y = element_blank(), 
           panel.border = element_blank(), 
           panel.grid.major = element_blank(), 
           panel.grid.minor = element_blank(), 
           axis.line = element_line(colour = "black", 
                                    size = rel(1)), 
           legend.key = element_blank(), 
           legend.title = element_blank(), 
           panel.background = element_rect(fill = "white", 
                                           colour = NA),
           strip.background = element_rect(fill = "white", colour = "black", 
                                           size = rel(2)), complete = TRUE)


c
```

```{r}
# Put together in figure
# Title "Environment type, rather than treatment variables, determines microbial composition over in bioreactor samples"
p <- plot_grid(a, b, c,
               align = 'h',
               rows = 1)
p

title <- ggdraw() + 
  draw_label(
    "Source environment, rather than treatment variables,\ndetermines microbial composition in bioreactor samples",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(
    # add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 45)
  )

plot_grid(
  title, p,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.1, 1)
)

ggsave(file.path(figs.dir, 'Beta diversity.pdf'),
                 width = 10, height = 4)
```


### PCoA 

Want to check the above given how astoundingly clean/striking the plot is.  Think this could be secondary to the timeseries nature of the data, but want to be sure.

## Glom to family