---
title: "Beta diversity"
author: "Brianna Petrone"
date: "3/31/2021"
output: html_document
---

# Setup

```{r}
# Install packages

# If not yet installed, need to install driver (for CLR transform) as follows
# devtools::install_github("jsilve24/driver")
```

```{r include=FALSE, echo=FALSE}
# Load libraries
library(driver)
library(phyloseq)
library(tidyverse) # For data wrangling tools
```

```{r}
# Set paths
data.dir <- '/data/project_data/argonne_data' # Data
scratch.dir <- '/home/guest/scratch/plastics' # Scratch
group.dir <- '/sharedspace/PlasticPeople' # Group
figs.dir <- file.path(scratch.dir, 'figures') # Figures
```

# Read in data

```{r}
# Phyloseq object
ps <- readRDS(file.path(group.dir, 'plastics.rds'))
ps
```

# Preprocessing

## Transform

```{r}
# Plan ahead
# Want to add pseudocount before CLR-transforming; easiest to do this now, while data are still interpreted as counts

# Pseudocount of 1 for compatibility with log transform
otu_table(ps) <- otu_table(ps) + 1 

# Check it worked
asvtab <- otu_table(ps)@.Data
min(asvtab)
```

```{r}
# Transform to relative abundance
ps.ra <- transform_sample_counts(ps, function(x){x/sum(x)})
```

## Filter

Want to do a thoughtful filtering step here.  For streamgraphs tried glomming to family level, then only keeping taxa with an abundance of 1% in at least 4 samples (because we might expect each location to be reasonably different, but is tested in 4 conditions at each timepoint).  

See how stringent this is, and how many more taxa are included if relaxed (perhaps to 2 samples)?

At the ASV level, we may want to relax this further (say to 0.01%)?

```{r}
# 0.01% in at least 4 samples
min_fraction = 0.0001
min_samples = 4
filter_taxa(ps.ra, 
            function(x) sum(x >= min_fraction) >= min_samples, prune=TRUE)
```

This reduces the number of taxa by `r 100*(1-(3547/ntaxa(ps)))`%.

```{r}
# 0.01% in at least 2 samples
min_fraction = 0.0001
min_samples = 2
filter_taxa(ps.ra, 
            function(x) sum(x >= min_fraction) >= min_samples, prune=TRUE)
```

This reduces the number of taxa by `r 100*(1-(6069/ntaxa(ps)))`%.

```{r}
# 0.1% in at least 1 samples
min_fraction = 0.001
min_samples = 1
ps.ra.filt <- 
     filter_taxa(ps.ra, 
                 function(x) sum(x >= min_fraction) >= min_samples, prune=TRUE)
ps.ra.filt
```

This reduces the number of taxa by `r 100*(1-(2933/ntaxa(ps)))`%.

Choosing the third filter condition for now.  I think a tenth of a percent is a meaningful relative abundance, and I like permitting keeping the threshold at that amount in 1 sample because there's so little replication in this dataset.

# Ordinate

## By ASV

### PCA

#### CLR transform input

```{r}
# Pseudocount has already been added, pre-transformation, so no need to do here

# Get ASV table
asvtab <- otu_table(ps.ra.filt)@.Data

# Transform
asvtab.clr <- clr(asvtab)

# Replace names, but remember these aren't interpretable as strict ASVs any longer
colnames(asvtab.clr) <- colnames(asvtab)

# Make updated phyloseq object
ps.clr <- phyloseq(otu_table(asvtab.clr, taxa_are_rows = FALSE),
                   sample_data(sample_data(ps)),
                   tax_table(tax_table(ps)))
```

#### Run

```{r}
# Run PCA
pca <- prcomp(asvtab.clr)
```

#### Plot

```{r}
# Get sample data
samdf <- 
    sample_data(ps.clr) %>% 
    as(., 'data.frame') 

# Add back sample data to PCA results
pca.df <- 
     data.frame(pca$x) %>% 
     rownames_to_column(var = 'sample')
pca.df <- left_join(pca.df, rownames_to_column(samdf, var = 'sample'))

# Figure out axis limits to customize scale of data
# Find the largest magnitude datapoint in the first 2 PCs, then add 5% to it for 
# plotting "room"
limit <- max(abs(pca.df[, c('PC1','PC2')])) +
     0.05*(max(abs(pca.df[, c('PC1','PC2')])))

# % variance explained
eigs <- pca$sdev^2
ve.pc1 <- as.character(100*round(eigs[1] / sum(eigs), 3))
ve.pc2 <- as.character(100*round(eigs[2] / sum(eigs), 3))
```

```{r}
# Plot by location
pca.plot <- 
     ggplot(pca.df, aes(PC1, PC2, color = Location)) +
     geom_point() +
     coord_equal() +
     xlim(-limit, limit) + ylim(-limit, limit) + 
     labs(x = paste0(' PC1 (', ve.pc1, '%)'),
          y = paste0(' PC1 (', ve.pc2, '%)'),
          color = 'Sample source') + 
     theme_bw()

pca.plot
```

```{r}
# Plot by timecourse
pca.plot <- 
     ggplot(pca.df, aes(PC1, PC2, color = Week)) +
     geom_point() +
     coord_equal() +
     xlim(-limit, limit) + ylim(-limit, limit) + 
     labs(x = paste0(' PC1 (', ve.pc1, '%)'),
          y = paste0(' PC1 (', ve.pc2, '%)'),
          color = 'Bioreactor week') + 
     theme_bw()

pca.plot
```

```{r}
# Plot by treatment
pca.plot <- 
     ggplot(pca.df, aes(PC1, PC2, color = Condition)) +
     geom_point() +
     coord_equal() +
     xlim(-limit, limit) + ylim(-limit, limit) + 
     labs(x = paste0(' PC1 (', ve.pc1, '%)'),
          y = paste0(' PC1 (', ve.pc2, '%)'),
          color = 'Treatment condition') + 
     theme_bw()

pca.plot
```

### PCoA 

## Glom to family