---
title: "LEfSe"
author: "Katherine Mueller"
date: "3/22/2021"
output: html_document
---

###LEfSe Analysis

##Setup
#Load Libraries
```{r}
library(readr)
library(phyloseq)
library(tibble)
library(dplyr)
library(ggplot2)
library(microbiomeMarker)
```

# Define Paths
```{r}
plastics.rds = "/sharedspace/PlasticPeople/plastics.rds"
```

# Check Phyloseq RDS
Load your phyloseq RDS and `print` it just to be sure things look sane
```{r}
plastics.ps = read_rds(plastics.rds)
print(plastics.ps)
```

```{r}
sample_variables(plastics.ps)
```

## Styrene Monomer
We noticed from the alpha diversity analysis that in the RC and AS locations, the growth conditions with styrene monomer tracked together separately from the other two growth conditions. Thus, we want to use LEfSe to see if there are differences in the taxa associated with styrene monomer.

In order to compare styrene containing versus styrene free media, I will need to create a new sample variable.
```{r}
metadata <- plastics.ps %>%
     sample_data() %>%
     as.data.frame()

conditions <- metadata$Condition
monomer <- ifelse(conditions %in% c("PSS", "S"), "has_monomer", "no_monomer")

sample_data(plastics.ps)$monomer_presence <- monomer
```

```{r}
sample_variables(plastics.ps)
```

Since we noticed this difference towards the end of the growth period, we should look specifically at the differences between groups at on weeks 7 and 12.
```{r}
endpoint.ps <- plastics.ps %>%
     subset_samples(Week %in% c(7,12))
print(endpoint.ps)
```

Since we want to find bacterial groups that could be tested in vitro for growth in plastic-containing media, it would be most convenient to identify specific species. Thus, I'll try filtering for just ASVs which were identifiable down to the species level.
```{r}
good.genus.ps <- endpoint.ps %>%
     subset_taxa(!is.na(Genus))
print(good.genus.ps)
```

#Overall PSS/S versus Ctrl/PS
First, I want to see if there are any differences in taxa between media with and without styrene, regardless of the location that samples were originally taken from.
```{r}
overall.lef <- lefse(endpoint.ps,
                     class = "monomer_presence",
                     lda_cutoff = 4,
                     correct = "1")
plot_ef_bar(overall.lef)
```
```{r}
#after filtering for ASVs identified at the species level
overall.lef2 <- lefse(good.genus.ps,
                     class = "monomer_presence",
                     lda_cutoff = 4,
                     correct = "1")
plot_ef_bar(overall.lef2, label_level = 0)
```


#PSS/S versus Ctrl/PS in RC and AS locations
Now, I want to compare media with and without styrene, accounting for the fact that we found the differences in alpha diversity in the samples from the RC and AS locations.
```{r}
RCAS.endpoint.ps <- endpoint.ps %>%
     subset_samples(Location %in% c("RC", "AS"))
print(RCAS.ps)
```

```{r}
RCAS.lef <- lefse(RCAS.endpoint.ps,
                     class = "monomer_presence",
                     lda_cutoff = 4,
                     multicls_strat = TRUE)
plot_ef_bar(RCAS.lef)
```
```{r}
RCAS.goodgenus.ps <- good.genus.ps %>%
     subset_samples(Location %in% c("RC", "AS"))
print(RCAS.goodspecies.ps)
```

```{r}
RCAS.lef2 <- lefse(RCAS.goodgenus.ps,
                     class = "monomer_presence",
                     lda_cutoff = 4,
                     multicls_strat = TRUE)
plot_ef_bar(RCAS.lef2, label_level = 0)
```

## Simplify LEfSe outputs
All of these outputs have been pretty messy to look at, even with the requirement for an LD score of 4. This is because the output includes taxonomic groups higher than genus. While this could be useful, we don't really need these higher levels while we're focusing on the genus/species of each ASV. We can keep Family around for ease of identification
```{r}
#Combine the genus and species identifications
taxa <- tax_table(good.genus.ps) %>%
     as.data.frame()
taxa$Genus_new <- paste(taxa$Genus, taxa$Species, sep = "_")

#Get rid of everything but the genus/species combo
new_taxa <- taxa %>%
     select(Genus_new)
colnames(new_taxa)[1] <- "Genus"

#Replace the taxa table in the phyloseq object with the new genus/species only table
clean.plastics.ps <- good.genus.ps
tax_table(clean.plastics.ps) <- as.matrix(new_taxa)
```

Overall PSS/S versus Ctrl/PS
```{r}
clean.overall.lef <- lefse(clean.plastics.ps,
                     class = "monomer_presence",
                     lda_cutoff = 3,
                     correct = "1")
plot_ef_bar(clean.overall.lef)
```

Just the RC and AS locations
```{r}
#Filter for locations of interest
RCAS.clean.ps <- clean.plastics.ps %>%
     subset_samples(Location %in% c("RC", "AS"))
print(RCAS.clean.ps)
```
```{r}
#Perform lefse analysis
clean.RCAS.lef <- lefse(RCAS.clean.ps,
                        class = "monomer_presence",
                        lda_cutoff = 3,
                        correct = "1")
plot_ef_bar(clean.RCAS.lef) +
     scale_fill_manual(values = c("has_monomer" = "blue", "no_monomer" = "grey"))
```