---
title: "Raw reads to phyloseq object"
output:
  html_document:
    df_print: paged
---

```{r setup}
library(Biostrings) # for reverse complementing barcodes
library(dada2)
library(fs) # For dir_create, dir_delete, dir_exists
library(phyloseq)
library(tidyverse)
```

## Setup

```{r}
# Set variables for R
# Directories
data.dir <- '/data/project_data/argonne_data' # Data
scratch.dir <- '/home/guest/scratch/plastics' # Scratch
demux.dir <- file.path(scratch.dir, 'demux') # Demultiplexed files
# Remove old demultiplexed files before starting again
if (dir_exists(demux.dir)) {dir_delete(demux.dir)}
dir_create(demux.dir)
# Make fresh demux directory
if (!dir.exists(demux.dir)) {dir.create(demux.dir)} 
group.dir <- '/sharedspace/PlasticPeople' # Group

# Files
map.file <- file.path(data.dir, "200114_McCumber_16SFW_AS_200110_corrected_final.txt")
rc.barcode.table <- file.path(scratch.dir, "fastq_multx_input_rc.txt")
```

```{r}
# Set variables for bash
# Directories
Sys.setenv(DATA_DIR = data.dir)
Sys.setenv(SCRATCH_DIR = scratch.dir)
Sys.setenv(DEMUX_DIR = demux.dir)
Sys.setenv(MAP_FILE = file.path(data.dir, "200114_McCumber_16SFW_AS_200110_corrected_final.txt"))

# Files for demultiplexing
# Make pointer to text file to hold subset of mapping file for fastq-multx
Sys.setenv(BARCODE_TABLE = file.path(scratch.dir, "fastq_multx_input.txt"))
Sys.setenv(RC_BARCODE_TABLE = file.path(scratch.dir, "fastq_multx_input_rc.txt"))
Sys.setenv(R1_FASTQ = file.path(data.dir, "Undetermined_S0_L001_R1_001.fastq.gz"))
Sys.setenv(R2_FASTQ = file.path(data.dir, "Undetermined_S0_L001_R2_001.fastq.gz"))
Sys.setenv(BARCODE_FASTQ = file.path(data.dir, "Undetermined_S0_L001_I1_001.fastq.gz"))
```

```{bash}
# Check data integrity
cd $DATA_DIR
md5sum -c md5_checksum_compressed_fastqs.txt
```
## Demultiplex

```{bash}
# Check structure of map file
head $MAP_FILE
```
### Forward barcodes
```{bash}
# Cut first two columns (sample ID and barcode) and place in barcode table to past to fastq-multx
set -u
cut --fields 1,2 $MAP_FILE > $BARCODE_TABLE

# Check to make sure looks as expected
head $BARCODE_TABLE
```

```{bash}
# Run fastq-multx
set -u
fastq-multx -m 3 -d 2 -x -B $BARCODE_TABLE \
  $BARCODE_FASTQ \
  $R1_FASTQ \
  $R2_FASTQ \
  -o $DEMUX_DIR/%_I1.fastq.gz \
  -o $DEMUX_DIR/%.forward.fastq.gz \
  -o $DEMUX_DIR/%.reverse.fastq.gz
```
11,712,260 total reads
2,271,549 unmatched reads
129,505 skipped because of edit distance > 2. 

### Reverse-complemented barcodes

Note: tried this once, determined forward orientation to be better:  Exclude from future runs.

```{r eval = FALSE}
# Remove output from the previous demultiplexing step
if (dir_exists(demux.dir)) {dir_delete(demux.dir)}
dir_create(demux.dir)
```

```{r eval = FALSE}
# Reverse-complement the provided barcodes, save them in fastq-multx-compatible format
read_tsv(map.file) %>% 
  select(Sample = "#SampleID", BarcodeSequence) %>%          
  deframe %>% # Convert to named vector (expected input for DNAStringSet constructor)
  DNAStringSet %>% # Convert to DNAStringSet
  reverseComplement %>%   
  as.data.frame %>% 
  rownames_to_column %>% 
  write_delim(rc.barcode.table, delim="\t", col_names=FALSE)
```

```{bash eval = FALSE}
set -u
fastq-multx -m 3 -d 2 -x -B $RC_BARCODE_TABLE \
  $BARCODE_FASTQ \
  $R1_FASTQ \
  $R2_FASTQ \
  -o $DEMUX_DIR/%_I1.fastq.gz \
  -o $DEMUX_DIR/%.forward.fastq.gz \
  -o $DEMUX_DIR/%.reverse.fastq.gz
```

Total reads: 11,712,260
Unmatched: 10,749,151
Skipped because edit distance < 2: 256,285

### Select project samples

Need to separate our project's files from others in the dataset.

```{r}
metadata <- 
  read_tsv(map.file) %>% 
  select(Sample = "#SampleID", everything())
head(metadata)
```
```{r}
# Total number of samples on run
dim(metadata)

# Oours are those with info for week, location, and condition
dim(filter(metadata, !is.na(Condition)))
```

Does this line up with how many samples we expect?
4 sites x 4 conditions x 8 weeks (1-7, and 12) = 128-- great.

```{r}
# Trim metadata to these samples only
metadata <- filter(metadata, !is.na(Condition))

# Get their sample names (for pulling files)
samples <- metadata$Sample
```