---
title: "Streamgraph"
author: "Brianna Petrone"
output: html_notebook
---

## Setup

```{r}
# Install packages

# If not yet installed, need to install streamgraph as follows
# devtools::install_github("hrbrmstr/streamgraph")
```

```{r}
# Load libraries
library(htmlwidgets) # For saving interactive streamgraph output
library(streamgraph) # For streamgraphs
library(tidyverse) # For data wrangling tools
```

```{r}
# Set path variables
# Directories
data.dir <- '/data/project_data/argonne_data' # Data
scratch.dir <- '/home/guest/scratch/plastics' # Scratch
group.dir <- '/sharedspace/PlasticPeople' # Group
figs.dir <- file.path(scratch.dir, 'figures') # Figures
```

## Read in data

```{r}
# Phyloseq object
ps <- readRDS(file.path(group.dir, 'plastics.rds')) # Raw data
# ps <- readRDS(file.path(group.dir, 'plastics_imputed.rds')) # Raw, imputed data
# ps <- readRDS(file.path(group.dir, 'plastics_imputed_filt.rds')) # Filtered, imputed data
ps
```

### Describe

```{r}
# How many taxa?
ntaxa(ps)
```

```{r}
# What phylogenetic level are these taxa specified to?

# Get taxonomy table as data frame
taxtab <- data.frame(tax_table(ps)@.Data)

percent.id <- colSums(!is.na(taxtab))/nrow(taxtab)
percent.id
```

```{r}
# Convert to dataframe for easier plotting
percent.id.df <- 
     data.frame(pct = percent.id) %>% 
     rownames_to_column(var = 'level')

# Make taxonomic level a factor variable so it shows up in order on axis
percent.id.df$level <- factor(percent.id.df$level,
                              levels = percent.id.df$level)

# Quick plot to illustrate
ggplot(percent.id.df, aes(x = level, y = pct)) +
     geom_bar(stat = 'identity') +
     labs(y = 'ASVs assigned (%)') +
     theme_bw() +
     theme(axis.title.x = element_blank(),
           axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
ggsave(file.path(figs.dir, '/ASVs assigned to taxonomic level.png'),
       height = 3, width = 4)
```

```{r}
# How many unique entries at each level?
# (Use to guide visualization)

apply(taxtab, 2, function(x){length(unique(x))})
```
Wow, this is a lot of entries.  May make visualization tricky.  Let's try glomming to the highest level (phylum) and seeing how it looks.

## Transform

```{r}
ps.ra <- transform_sample_counts(ps, function(x){x/sum(x)})
```

### Glom 

```{r}
# ps.family <- tax_glom(ps.ra, taxrank = 'Family', NArm = FALSE)
```

## Filter
The other idea would be to filter low-abundance taxa to see if this helps at all.

```{r}
# Only keep taxa with an abundance of 1% in at least one sample
min_fraction = 0.01
min_samples = 1

# On glommed data
# ps.family.filt = filter_taxa(ps.family, 
#                              function(x) sum(x >= min_fraction) >= min_samples,
#                              prune=TRUE)

# On raw data
ps.ra.filt = filter_taxa(ps.ra,
                         function(x) sum(x >= min_fraction) >= min_samples,
                         prune=TRUE)
```

## Streamgraph

From (R Graph Gallery)[https://www.r-graph-gallery.com/154-basic-interactive-streamgraph-2.html]:
Your input needs 3 columns:

* a numeric variable for the X axis. Usually it provides time.
* a numeric variable for the Y axis: this is the variable we are studying
* a categorical variable: each group will be plotted as a distinct shape

```{r}
melt <- psmelt(ps.ra.filt)

# Make joined name for each OTU
melt <- mutate(melt,
               name = paste(Kingdom, Phylum, Class, Order, Family, Genus,
                            Species, sep = ';'))

# Convert to factor
melt$name <- factor(melt$name)
```

```{r}
n_distinct(melt$OTU)
n_distinct(melt$name)
```
Nice; this reduces further.

```{r}
melt <- 
     melt %>% 
     group_by(Week, Location, Condition, name) %>% 
     summarize(Abundance = sum(Abundance))
```

```{r}
# Basic stream graph: just give the 3 arguments
pp <- 
     melt %>% 
     filter(Location == 'RC' & Condition == 'Ctrl') %>% 
     streamgraph(key=name, value=Abundance, date=Week, scale = 'continuous',
                 height="300px", width="400px")

pp
```

### Make all 16

```{r}
locations <- unique(sample_data(ps)$Location)
conditions <- unique(sample_data(ps)$Condition)

for (loc in locations){
     for (cond in conditions){
          # Make streamgraph
          pp <- 
               melt %>% 
               filter(Location == loc & Condition == cond) %>% 
               streamgraph(key=name, 
                           value=Abundance, date=Week, scale = 'continuous',
                           height="300px", width="400px")
          
          # Save
          saveWidget(pp, file=file.path(figs.dir,
                                        'streamgraphs_filt',
                                        paste(loc, cond, 'streamgraph.html',
                                              sep = '_')))
     }
}

```

